Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-02 talk/icooolps2009/paper.tex
minor clarification and some notes

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand{\commentout}[1]{}

@@

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers. XXX

@@

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that was being interpreted so far.

@@

XXX write somewhere here on which level the RPython tracer operates

@@

This trace will then be turned into machine code. Note that the machine code
loop is by itself infinite and can only be left via a guard failure. Also note
\texttt{f} was inlined into the loop and how the common \texttt{else} case was
turned into machine code, while the other one is implemented via a guard
failure. The variables in square brackets after the guards are the state that
the interpreter will get when the guard fails.

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

>>>>

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand{\commentout}[1]{}

@@

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers. XXX [fijal] cite andreas gal paper?

@@

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the interpreted program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
starting from the second iteration of the loop,
as it represents exactly the loop that was being interpreted so far.

@@

\fijal{RPython tracer operates on a level which is slightly higher level than
of the underlaying C. To be more precise: it operates on exception-transformed,
non-gc-transformed graphs with an extra bit of abstraction inversion that
finds out operations on loops and dicts, not sure how to express this}

@@

\fijal{Following paragraph is more confusing than helpful. The trace contains
all operations performed, including full inlining of functions called.
The loop will loop over the path in the interpreter, which corresponds to
path taken when tracing (which we assume is a common case).
Later, which is outside of the scope of this paper, less common cases
might be turned into assembler
code as well, producing bridges, if they're common enough. Note that this
loop is infinite, which means the only way to exit it is via guard failure}
This trace will then be turned into machine code. Note that the machine code
loop is by itself infinite and can only be left via a guard failure. Also note
\texttt{f} was inlined into the loop and how the common \texttt{else} case was
turned into machine code, while the other one is implemented via a guard
failure. The variables in square brackets after the guards are the state that
the interpreter will get when the guard fails.

@@

\fijal{I'm lost, will continue later}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
